A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-07-10 paper/theor-experiment.tex
rearranging AWA section

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization} \label{awaproblem}
Next we apply the optimization framework to the high charge beam line at the AWA. 
The purpose of the Argonne Wakefield Accelerator (AWA) optimization is to produce beams of electrons meeting 
design specifications including number of particles (charge), energy, and particle distribution (characterized by transverse beam sizes and energy spread).
As shown in Fig.  \ref{awa-pic}, this beam line consists of an rf photocathode gun, 
two solenoids, six linear accelerating cavities, 
followed by four quadrupoles and a stripline kicker. 
The charge of interest, 40 nC, is needed for Two Beam 
Acceleration (TBA) \cite{gai_power_jing_2012} experiments performed in the past 
at AWA. Similar experiments will take place in the future and motivate this work. 
Prior experimental results were limited by beam size as the beam passed through small aperture 
wakefield structures located after the quadrupoles. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize several parameters in the gun and quadrupoles leading into the 
TBA section of the beam line.
%\vspace{-1em}

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig1) at (0,0)
	{\includegraphics[width=\textwidth]{Report/awa-drawing}};
	\node[fill=white, inner sep=2pt] (txt2) at (125,10) {$s_2$};
	\node[fill=white, inner sep=2pt] (txt2) at (112,8) {$s_1$};
	\node[fill=white, inner sep=2pt] (txt2) at (115,30) {Kicker};
	\draw [blue, ->, line width=1] (120,30) -- (120, 23);	
	\end{tikzpicture}
	\caption{Side view of the high charge linac at the AWA. 
		All hardware in this drawing is currently installed.}
	\label{awa-linac}
\end{figure}

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig2) at (0,0)
	{\includegraphics[width=\textwidth]{Report/tba_dogleg}};
	\node[fill=white, inner sep=2pt] (txt2) at (2,35) {$s_1$};
	\node[fill=white, inner sep=2pt] (txt2) at (14,35) {$s_2$};
	\node[fill=white, inner sep=2pt] (txt2) at (33,31) {$s_3$};
	\node[fill=white, inner sep=2pt] (txt2) at (24,50) {Septum};
	\draw [blue, ->, line width=1] (30,50) -- (30, 42);	
	\node[fill=white, inner sep=2pt] (txt2) at (3,50) {Kicker};
	\draw [blue, ->, line width=1] (9.5,50) -- (9.5, 43);
	\end{tikzpicture}
	\caption{Continuation of the high charge beam line layout at the AWA, top view. 
		This is the proposed two beam acceleration section.}
	\label{awa-tba}
\end{figure}

%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, 
along with three constraints.
The objectives include transverse and longitudinal beam size, 
transverse momentum and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_1=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-pic}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.
\nrnote{Maybe eqs 5.10-5.21 and code example should move after time step scan and hyper parameters}

\begin{align}
\text{min}  \quad & \text{rms}_{x}\left(s = s_3\right), \quad \text{rms}_{y}\left(s = s_3\right) \label{eq:awa:p1}\\
& \text{rms}_{px}\left(s = s_3\right), \quad \text{rms}_{py}\left(s = s_3\right), \label{eq:awa:p2}\\
& \text{rms}_{s}\left(s = s_3\right), \quad dE \label{eq:awa:p4} \\
\text{subject to} \quad & rms_x(s=s_1) < 0.1 \, [m]\label{eq:awa:c1}\\
\quad & rms_y(s=s_1) < 0.1\, [m]\label{eq:awa:c2}\\
\quad & |rms_y(s=s_1) - rms_x(s=s_1) | < 0.005 \, [m]\label{eq:awa:c3}\\
\quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & R_x = R_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{gun}} =-20^\circ \label{eq:awa:gphidvar}\\
\quad & \phi_{\text{linac}} =-20^\circ \label{eq:awa:lastdvar}
\end{align}

The first four objectives, Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p2}),
minimize the transverse ($rms_{x,y}$) beam size and transverse momentum ($rms_{px,py}$)
at the location of interest in the beam line ($s_1$).
Where $s_1$ is the longitudinal entrance of the fifth quadrupole. 
Minimizing the beam size at this location is essential to 
to preventing loss of particles by scraping.
Preventing scraping ensures better transmission through the wakefield structures downstream. 
Less divergence in the beam (lower transverse momentum) 
reduces growth of transverse beam size during transport in bending elements, 
which also helps with transmission downstream. 
The next two objectives (\ref{eq:awa:p2}) minimize the 
longitudinal beam size ($rms_s$), and energy spread (dE) at $s_1$.
This helps reduce longitudinal beam size growth in bending elements.

@@

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNLs.
This in combination with the number of nodes available 
allows for very large optimization jobs, like the AWA case.
Typical runs for this paper used 41 nodes, which corresponds to 2624 KNL cores 
and a generation size of 328 individuals.

\subsubsection{Time Step Scan} \label{awa:subsection:test}
A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size $16 \times 16 \times 32$ was chosen, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. The
average run time of each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes.
%A summary of the parameters used is listed in Table \ref{fidelity}.

\subsubsection{AWA Optimization Results}
With the time steps and hyper parameters set by the work in \ref{awa:subsection:test}, 
the full optimization problem described in \ref{awaproblem} was run for 200 generations.
The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers were in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources.

>>>>

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization} \label{awaproblem}
Next we apply the optimization framework to a high charge beam line
 at the Argonne Wakefield Accelerator (AWA) facility. 
The goal of this optimization is to produce beams of electrons that meet 
design specifications; this includes number of particles (charge), energy, and particle distribution (characterized by beam sizes and energy spread).
As shown in Fig.~\ref{awa-linac}, the installed portion of the the 
beam line consists of an rf photocathode gun, 
two solenoids, and six linear accelerating cavities
followed by four quadrupoles and a stripline kicker. 
The charge of interest, 40 nC, is needed for Two Beam 
Acceleration (TBA) \cite{gai_power_jing_2012,JING201872} 
experiments performed at AWA, which motivates this work. 
Prior experimental results were limited by beam size as the beam passed through small aperture 
wakefield structures located downstream. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize magnet strengths in the gun and quadrupoles leading into the 
TBA section of the beam line, shown in Fig.~\ref{awa-tba}.
The simulation model includes the linac, kicker, and septum. 
The optimization location is chosen as the entrance to the fifth quadrupole, $s_3$, on the dog leg.
Minimizing beam sizes here will enable capture and further focus with quadrupoles five and six before space charge effects dominate the beam. 
This will also enable cleaner transport through downstream elements.
%\vspace{-1em}

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig1) at (0,0)
	{\includegraphics[width=\textwidth]{Report/awa-drawing}};
	\node[fill=white, inner sep=2pt] (txt2) at (0,20) {Gun};
	\node[fill=white, inner sep=2pt] (txt2) at (20,20) {Accelerating Cavities};
	\node[fill=white, inner sep=2pt] (txt2) at (88,28) {Quadrupoles};
	\draw [blue, ->, line width=1] (99,28) -- (99, 21);
	\node[fill=white, inner sep=2pt] (txt2) at (125,10) {$s_2$};
	\node[fill=white, inner sep=2pt] (txt2) at (112,8) {$s_1$};
	\node[fill=white, inner sep=2pt] (txt2) at (115,30) {Kicker};
	\draw [blue, ->, line width=1] (120,30) -- (120, 23);	
	\end{tikzpicture}
	\caption{Side view of the high charge linac at the AWA. 
		All hardware in this drawing is currently installed.}
	\label{awa-linac}
\end{figure}

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig2) at (0,0)
	{\includegraphics[width=\textwidth]{Report/tba_dogleg}};
	\node[fill=white, inner sep=2pt] (txt2) at (2,35) {$s_1$};
	\node[fill=white, inner sep=2pt] (txt2) at (14,35) {$s_2$};
	\node[fill=white, inner sep=2pt] (txt2) at (33,31) {$s_3$};
	\node[fill=white, inner sep=2pt] (txt2) at (24,50) {Septum};
	\draw [blue, ->, line width=1] (30,50) -- (30, 42);	
	\node[fill=white, inner sep=2pt] (txt2) at (3,50) {Kicker};
	\draw [blue, ->, line width=1] (9.5,50) -- (9.5, 43);
	\end{tikzpicture}
	\caption{Continuation of the high charge beam line layout at the AWA, top view. 
		This is the proposed two beam acceleration section.}
	\label{awa-tba}
\end{figure}
%
%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, along with three constraints.
The objectives include transverse and longitudinal beam sizes, 
transverse momentum, and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_3=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-tba}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.
\nrnote{Moved final run set up and results to after time step scan and hyper parameters}

\subsection{Tests}
Before running a full scale optimization of the problem described in \ref{awaproblem}, 
several computational settings, and 
\subsubsection{Time Step Scan} \label{awa:subsection:test}
A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size $16 \times 16 \times 32$ was chosen, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. The
average run time of each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes.
%A summary of the parameters used is listed in Table \ref{fidelity}.

\subsubsection{AWA Optimization Problem}
\begin{align}
\text{min}  \quad & \text{rms}_{x}\left(s = s_3\right), \quad \text{rms}_{y}\left(s = s_3\right) \label{eq:awa:p1}\\
& \text{rms}_{px}\left(s = s_3\right), \quad \text{rms}_{py}\left(s = s_3\right), \label{eq:awa:p2}\\
& \text{rms}_{s}\left(s = s_3\right), \quad dE \label{eq:awa:p4} \\
\text{subject to} \quad & rms_x(s=s_1) < 0.1 \, [m]\label{eq:awa:c1}\\
\quad & rms_y(s=s_1) < 0.1\, [m]\label{eq:awa:c2}\\
\quad & |rms_y(s=s_1) - rms_x(s=s_1) | < 0.005 \, [m]\label{eq:awa:c3}\\
\quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & R_x = R_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{gun}} =-20^\circ \label{eq:awa:gphidvar}\\
\quad & \phi_{\text{linac}} =-20^\circ \label{eq:awa:lastdvar}
\end{align}

The first four objectives, Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p2}),
minimize the transverse ($rms_{x,y}$) beam size and transverse momentum ($rms_{px,py}$)
at the location of interest in the beam line ($s_3$).
%Where $s_3$ is the longitudinal entrance of the fifth quadrupole. 
Minimizing the beam size at this location is essential to 
to preventing loss of particles by scraping.
Preventing scraping ensures better transmission through the wakefield structures downstream. 
Less divergence in the beam (lower transverse momentum) 
reduces growth of transverse beam size during transport in bending elements, 
which also helps with transmission downstream. 
The next two objectives (\ref{eq:awa:p2}) minimize the 
longitudinal beam size ($rms_s$), and energy spread (dE) at $s_1$.
This helps reduce longitudinal beam size growth in bending elements.

@@

\subsubsection{AWA Optimization Results}
All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNLs.
This in combination with the number of nodes available 
allows for very large optimization jobs, like the AWA case.
Typical runs for this paper used 41 nodes, which corresponds to 2624 KNL cores 
and a generation size of 328 individuals.

With the time steps and hyper parameters set by the work in \ref{awa:subsection:test}, 
the full optimization problem described in \ref{awaproblem} was run for 200 generations.
The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers were in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources.
